---
title: "A short introduction to cross-network analysis with xnet"
author: "Meys Joris"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{xnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6, fig.height = 4,out.width = '49%',fig.align = 'center',
  collapse = TRUE,
  comment = "#>"
)
suppressMessages(library(xnet))
```

## Concepts and terms used in the package.

### Notation and naming of networks in the package

Networks exist in all forms and shapes. With `xnet` you can analyse
bipartite networks, i.e. networks that describe the link between two
sets of nodes. For example:

 * which proteins interact with eachother?
 * which goods are bought by which clients?
 * How many likes give twitter users to eachother's tweets?

These two sets can contain the same nodes (eg protein interaction) or
different nodes (eg goods bought by clients). When the two sets are 
the same, we call this a *homogenous* network. A network between two
different sets of nodes is called a *heterogenous* network.

The interactions are presented in a *adjacency matrix*, noted **Y**. 
The rows of **Y** represent one set of nodes, the columns the second.
Interactions can be measured on a continuous scale, indicating
how strong each interaction is. Often the adjacency matrix only contains
a few values: 1 for interaction, 0 for no interaction and possibly -1 
for an inverse interaction.

two-step kernel ridge regression ( function `tskrr()` ) predicts the values
in  the adjacency matrix based on similarities within the node sets,
calculated by using some form of a *kernel function*. This function takes
two nodes as input, and outputs a measure of similarity with specific
mathematical properties. The resulting *kernel matrix* has to be positive
definite for the method to work. In the package, these matrices are noted
**K** for the rows and - if applicable - **G** for the columns of **Y**. 

We'd like to refer to the [kernlab](https://cloud.r-project.org/package=kernlab) for a collection
of different kernel functions.

### Data in the package

For the illustrations, we use two different datasets.

#### Homogenous networks

The example dataset `proteinInteraction` originates from a publication
by [Yamanishi et al (2004)](https://doi.org/10.1093/bioinformatics/bth910).
It contains data on interaction between a subset of 769 proteins, and
consists of two objects:

 * the adjacency matrix `proteinInteraction` where 1 indicates an interaction between proteins
 * the kernel matrix `Kmat_y2h_sc` describing the similarity between
 the proteins.

#### Heterogenous networks

The dataset `drugtarget` serves as an example of a heterogenous network
and comes from a publication of [Yamanishi et al (2008)](https://doi.org/10.1093/bioinformatics/btn162). In order to get
a correct kernel matrix, we recalculated the kernel matrices as explained
in the vignette [Preparation of the example data](../doc/Preparation_example_data.html). 

The dataset exists of three objects

 * the adjacency matrix `drugTargetInteraction`
 * the kernel matrix for the targets `targetSim`
 * the kernel matrix for the drugs `drugSim`
 
The adjacency matrix indicates which protein targets interact with which
drugs, and the purpose is to predict new drug-target interactions.

## Fitting a two-step kernel ridge regression

### Heterogenous network

To fit a two-step kernel ridge regression, you use the function `tskrr()`. This function needs to get some tuning parameter(s)
`lambda`. You can choose to set 1 lambda for tuning **K**
and **G** using the same lambda value, or you can specify
a different lambda for **K** and **G**.

```{r fit a heterogenous model}

data(drugtarget)

drugmodel <- tskrr(y = drugTargetInteraction,
                   k = targetSim,
                   g = drugSim, 
                   lambda = c(0.01,0.1))

drugmodel
```

### Homogenous network

For homogenous networks you use the same function, but you don't specify
the **G** matrix. You also need only a single lambda:

```{r fit a homogenous model}
data(proteinInteraction)

proteinmodel <- tskrr(proteinInteraction,
                      k = Kmat_y2h_sc,
                      lambda = 0.01)

proteinmodel
```

### Extracting information from a model.

The model output itself tells you only little, apart from the dimensions,
the lambdas used and the labels found in the data. That information
can be extracted using a number of convenience functions.

```{r extract info from a model}
lambda(drugmodel)  # extract lambda values
lambda(proteinmodel)
dim(drugmodel) # extract the dimensions

protlabels <- labels(proteinmodel)
str(protlabels)
```

 * `lambda` returns a vector with the lambda values used.
 * `dim` returns the dimensions.
 * `labels` returns a list with two elements, `k` and `g`, containing
 the labels for the rows resp. the columns.
 
You can also use the functions `rownames()` and `colnames()` to extract
the labels.


### Information on the fit of the model

The functions `fitted()` and `predict()` can be used to extract the
fitted values. The latter also allows you to specify new kernel matrices
to predict for new nodes in the network. To get the residuals, you can
use the function `residuals()`. This is shown further in the document.

## Performing leave-one-out crossvalidation

### Settings for LOO

The most significant contribution of this package, are the various
shortcuts for leave-one-out crossvalidation (LOO-CV) described in 
[the paper by Stock et al, 2018](https://doi.org/10.1093/bib/bby095).
Generally LOO-CV removes a value, refits the model and predicts the
removed value based on this refit model. In this package you do this 
using the function `loo()`. The paper describes a number 
of different settings, which can be passed to the argument `exclusion`:

 * *interaction*: in this setting only the interaction between two nodes
 is removed from the adjacency matrix.
 * *row*: in this setting the entire row for that node is removed
 from the adjacency matrix. This boils down to removing a node from
 the **K** set.
 * *column*: in this setting the entire column for that node is removed
 from the adjacency matrix. This boils down to removing a node from the
 **G** set.
 * *both*: in this setting both rows and columns are removed, i.e. for
 every loo value the respective nodes are removed from both the **K**
 and the **G** set.
 
In some networks only information of interactions is available, so a 0
doesn't necessarily indicate "no interaction". It just indicates
"no knowledge" for an interaction. In those cases it makes more sense
to calculate the LOO values by replacing the interaction by 0 instead
of removing it. This is done by setting `replaceby0 = TRUE`.

```{r calculate loo values}
loo_drugs_interaction <- loo(drugmodel, exclusion = "interaction",
                       replaceby0 = TRUE)
loo_protein_both <- loo(proteinmodel, exclusion = "both")
```

In both cases the result is a matrix with the LOO values.

### Use LOO in other functions

There's quite a few functions that allow you to use LOO values instead
of predictions for calculations and optimizations. For example, you can
calculate residuals based on LOO values directly using the 
function `residuals()`:

```{r calculate loo residuals}
loo_resid <- residuals(drugmodel, method = "loo",
                       exclusion = "interaction",
                       replaceby0 = TRUE)
all.equal(loo_resid, 
          response(drugmodel) - loo_drugs_interaction )
```

Every other function that can use LOO values instead of predictions will
have the same two arguments `exclusion` and `replaceby0`.

## Looking at model output

The function provides a `plot()` function for looking at the model output.
This function can show you the fitted values, LOO values or the 
residuals. It also lets you construct dendrograms based on the **K** and
**G** matrices, so you have both the predictions and the similarity
information on the nodes in one plot. 

```{r plot a model, fig.show = 'hold'}
plot(drugmodel, main = "Drug Target interaction")
```

To plot LOO values, you set the argument `which`. As the protein model
is pretty extensive, we can remove the dendrogram and select a number
of proteins we want to inspect closer.

```{r plot the loo values}
plot(proteinmodel, dendro = "none", main = "Protein interaction - LOO",
     which = "loo", exclusion = "both",
     rows = rownames(proteinmodel)[10:20],
     cols = colnames(proteinmodel)[30:35])

```

If the colors don't suit you, you can set both the breaks used for
the color code and the color code itself. 

```{r plot different color code}
plot(drugmodel, which = "residuals", 
     col = rainbow(20),
     breaks = seq(-1,1,by=0.1))
```

## Tuning a model to find the best lambda.

In most cases you don't know how to set the `lambda` values for optimal
predictions. In order to find the best `lambda` values, the function
`tune()` allows you to do a grid search. This grid search can be
done in a number of ways:

 * by specifying actual values to be tested
 * by specifying the minimum and maximum lambda together with the
 number of values needed in every dimension. The function will create
 a grid that's even spaced on a log scale.

Tuning minimizes a loss function. Two loss functions are provided, i.e.
one based on mean squared error (`loss_mse`) and one based on the area
under the curve (`loss_auc`). But you can provide your own loss function
too if needed.

### Homogenous networks

Homogenous networks have a single lambda value, and should hence only
search in a single dimension. The following code tests 20 lambda values
between 0.001 and 10.

```{r tune a homogenous network}
proteintuned <- tune(proteinmodel,
                     lim = c(0.001,10),
                     ngrid = 20,
                     fun = loss_auc)
proteintuned
```

The returned object is a again a model object with the model fitted using
the best lambda value. It also contains extra information on the settings
of the tuning. You can extract the grid values as follows:
```{r get the grid values}
get_grid(proteintuned)
```
This returns a list with one or two elements, each element containing
the grid values for the respective kernel matrix.

You can also create a plot to visually inspect the tuning:
```{r plot grid}
plot_grid(proteintuned)
```

This object is also a tskrr model, so all the functions used above can
be used here as well. For example, we can use the same code as before
to inspect the LOO values of this tuned model:

```{r residuals tuned model}
plot(proteintuned, dendro = "none", main = "Protein interaction - LOO",
     which = "loo", exclusion = "both",
     rows = rownames(proteinmodel)[10:20],
     cols = colnames(proteinmodel)[30:35])
```

### Heterogenous networks

For heterogenous networks, the tuning works the same way. Standard, the
function `tune()` performs a two-dimensional grid search. To do a
one-dimensional grid search (i.e. use the same lambda for **K** and 
**G**), you set the argument `onedim = TRUE`.

```{r}
drugtuned1d <- tune(drugmodel,
                    lim = c(0.001,10),
                    ngrid = 20,
                    fun = loss_auc,
                    onedim = TRUE)
plot_grid(drugtuned1d, main = "1D search")
```

When performing a two-dimensional grid search, you can specify different
limits and grid values or lambda values for both dimensions. You do this
by passing a list with 2 elements for the respective arguments.

```{r tune 2d model}
drugtuned2d <- tune(drugmodel,
                     lim = list(k = c(0.001,10), g = c(0.0001,10)),
                     ngrid = list(k = 20, g = 10),
                     fun = loss_auc)
```

the `plot_grid()` function
will give you a heatmap indicating where the optimal lambda values are
found:

```{r plot grid 2d model}
plot_grid(drugtuned2d, main = "2D search")
```

As before, you can use the function `lambda()` to get to the best
lambda values. 

```{r}
lambda(drugtuned1d)
lambda(drugtuned2d)
```

It seems like a one-dimensional grid search gives quite different optimal
lambda values. To get more information on the loss values, you can use
the function `get_loss_values()`. This allows you to examine the actual
improvement for every lambda value. The output is always a matrix, and
in the case of a 1D search it's a matrix with one column. Combining
these values with the lambda grid, shows that the the difference between
a lambda value of around 0.20 and around 0.34 is very small. This is
also obvious from the grid plots shown above.

```{r}
cbind(
  loss = get_loss_values(drugtuned1d)[,1],
  lambda = get_grid(drugtuned1d)$k
)[10:15,]

```

## Predicting new values 

In order to predict new values, you need information on the outcome of
the kernel functions for the combination of the new values and those
used to train the model. Depending on which information you have, you can
do different predictions. To illustrate this, let's split up the data
for the drugsmodel.

```{r reorder the data}
idk_test <- c(5,10,15,20,25)
idg_test <- c(2,4,6,8,10)

drugInteraction_train <- drugTargetInteraction[-idk_test, -idg_test]
target_train <- targetSim[-idk_test, -idk_test]
drug_train <- drugSim[-idg_test, -idg_test]

target_test <- targetSim[idk_test, -idk_test]
drug_test <- drugSim[idg_test, -idg_test]
```

So the following drugs and targets are removed from the training data
and will be used for predictions later:
```{r}
rownames(target_test)
colnames(drug_test)
```

We can now train the data using `tune()` just like we would use `tskrr()`

```{r train the model}
trained <- tune(drugInteraction_train,
                k = target_train,
                g = drug_train,
                ngrid = 30)
```

### Predict for new K-nodes

In order to predict the interaction between new targets and the drugs
in the model, we need to pass the kernel values for the similarities
between the new targets and the ones in the model. The `predict()`
function will select the correct **G** matrix for calculating the 
predictions.

```{r}
Newtargets <- predict(trained, k = target_test)
Newtargets[, 1:5]
```

### Predict for new G-nodes

If you want to predict for new drugs, you need the kernel values for
the similarities between new drugs and the drugs trained in the model.

```{r}
Newdrugs <- predict(trained, g = drug_test)
Newdrugs[1:5, ]
```

### Predict for new K and G nodes

You can combine both kernel matrices used above to get predictions about
the interaction between new drugs and new targets:

```{r}
Newdrugtarget <- predict(trained, k=target_test, g=drug_test)
Newdrugtarget
```

## Impute new values based on a tskrr model

Sometimes you have missing values in a adjacency matrix. These missing
values can be imputed based on a simple algorithm:

 1. replace the missing values by a start value
 2. fit a tskrr model with the added values
 3. replace the missing values with the predictions of that model
 4. repeat until the imputed values converge (i.e. the difference with
 the previous run falls below a tolerance value)
 
Apart from the usual arguments of `tskrr`, you can give additional
parameters to the function `impute_tskrr`. The most important ones are

 * `niter`: the maximum number of iterations
 * `tol`: the tolerance, i.e. the minimal sum of squared differences 
 between iteration steps to keep the algorithm going
 * `verbose`: setting this to 1 or 2 gives additional info on the
 algorithm performance.
 
So let's construct a dataset with missing values:

```{r create missing values}
drugTargetMissing <- drugTargetInteraction
idmissing <- c(10,20,30,40,50,60)
drugTargetMissing[idmissing] <- NA
```

Now we can try to impute these values. The outcome is again a
tskrr model.

```{r}
imputed <- impute_tskrr(drugTargetMissing,
                        k = targetSim,
                        g = drugSim,
                        verbose = TRUE)
plot(imputed, dendro = "none")
```
To extract information on the imputation, you have a few convenience
functions to your disposal:

 * `has_imputed_values()` tells you whether the model contains imputed values
 * `is_imputed()` returns a logical matrix where `TRUE` indicates an
 imputed value
 * `which_imputed()` returns an integer vector with the positions of the
 imputed values. Note that these positions are vector positions, i.e.
 they give the position in a single dimension (according to how a 
 matrix is stored internally in R.)
 
```{r}
has_imputed_values(imputed)
which_imputed(imputed)

# Extract only the imputed values
id <- is_imputed(imputed)
predict(imputed)[id]
```

You can use this information to plot the imputed values in context:
```{r}
rowid <- rowSums(id) > 0
colid <- colSums(id) > 0
plot(imputed, rows = rowid, cols = colid)
```