---
title: "S4 class structure of the xnet package"
author: "Meys Joris"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{xnet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(xnet)

slotsToRmd <- function(class){
  slots <- getSlots(class)
  
  txt <- paste(" *",names(slots), ": object of type", slots)
  cat(txt, sep = "\n")
}
```

This document defines the S4 class structure used in the `xnet` package.

## Virtual classes

The `xnet` package has three virtual classes that each define a 
number of slots necessary for that specific type of model:

 1. the class `tskrr` for general two step kernel ridge regressions
 2. the class `tskrrTune` for tuned two step kernel ridge regressions
 3. the class `tskrrImpute` for two step kernel ridge regressions with
 imputed data.
 
Each of these classes defines the necessary slots for that specific 
type of action. The actual classes returned by the functions `tskrr()`,
`tune()` and `impute()` inherit from (a combination of) these virtual
classes.

## Actual classes

### Inheritance from tskrr

After using the function `tskrr()`, one of the following classes is
returned:

  * `tskrrHomogenous` : for homogenous networks
  * `tskrrHeterogenous` : for heterogenous networks
     
These classes have similar slots, but the homogenous models don't need
information on the column kernel matrix. The slots are listed below.

### Slots defined by tskrrHomogenous

```{r Create slots tskrrHomogenous, results='asis', echo = FALSE}
slotsToRmd("tskrrHomogenous")
```

### Slots defined by tskrrHeterogenous

```{r Create slots tskrrHeterogenous, results='asis', echo = FALSE}
slotsToRmd("tskrrHeterogenous")
```

Both classes inherit directly from the class `tskrr`. But these 
classes also function as parent classes from which `tune()` related
and `impute()` related classes inherit.

### Inheritance from tskrrTune

When using the function `tune()`, you get one of the following classes:

  * `tskrrTuneHomogenous` : for tuned homogenous networks. Inherits also from `tskrrHomogenous`.
  * `tskrrTuneHeterogenous` : for tuned heterogenous networks. Inherits also from `tskrrHeterogenous`.
     
Apart from the slots of the respective `tskrr` class, the inheritance
from `tskrrTune` adds the following slots:

```{r Create slots tskrrTune, results='asis', echo = FALSE}
slotsToRmd("tskrrTune")
```

     
### Inheritance from tskrrImpute

When using the function `impute()`, you get one of the following 
classes:

  * `tskrrHomogenousImpute` : for homogenous networks with imputed data.
  * `tskrrHeterogenousImpute` : for heterogenous networks with imputed data.

Apart from the slots of the respective `tskrr` class, the inheritance
from `tskrrImpute` adds the following slots:

```{r Create slots tskrrImpute, results='asis', echo = FALSE}
slotsToRmd("tskrrImpute")
```

