---
title: "Preparation of the example data"
author: "Joris Meys"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparation of the example data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The example data used in this package was originally published by
[Yamanishi et al, 2008](https://doi.org/10.1093/bioinformatics/btn162). They 
used [the KEGG data base](www.kegg.jp/) to get information drug-target
interaction for different groups of enzymes. We used their supplementary
material as a basis for the example data provided tothe package. Their
supplementary datasets can be downloaded from [here](web.kuicr.kyoto-u.ac.jp/supp/yoshi/drugtarget/)

## Obtaining the original data

The original adjacency matrix and similarity of the targets
were downloaded from that website using the following code:

```{r getFiles, eval = FALSE}
adjAddress <- "http://web.kuicr.kyoto-u.ac.jp/supp/yoshi/drugtarget/nr_admat_dgc.txt"

targetAddress <- "http://web.kuicr.kyoto-u.ac.jp/supp/yoshi/drugtarget/nr_simmat_dg.txt"

drugTargetInteraction <- as.matrix(
  read.table(adjAddress, header = TRUE, row.names = 1, sep = "\t")
)
targetSim <- as.matrix(
  read.table(targetAddress, header =TRUE, row.names = 1, sep = "\t")
)
```

This data was used as is from the website.

## Processing the drug similarities

In the original paper the authors relied on the SIMCOMP algorithm, but this
method returns non-symmetric matrices and hence the original data cannot
be used in a meaningful way for a two-step kernel ridge regression. Hence we decided to recreate the similarities between the different drugs,
this time using the algorithms provided in the [fmcsR package]
(https://bioconductor.org/packages/release/bioc/html/fmcsR.html).

TO BE COMPLETED
